A state space model for characterizing the trajectory dynamics of replay

### 2020-06-08 Article.tex

\subsection*{The model finds replays with non-constant speed dynamics}
After validating our model on a single SWR, we then applied our decoding algorithm to hippocampal recordings from 10 animals performing the W-track spatial alternation task (tetrodes = [10, 24], brain areas = [CA1, CA2, CA3]; some data previously used in \cite{KarlssonAwakereplayremote2009}). We observed many replays that were classified as continuous throughout the entirety of SWR and likely would have been identified by the typical Bayesian Decoder (Figure 2-Supplemental 2A-C). However, we also observed replays that did not have this structure, but could be interpreted in terms of our model. For example, we observed trajectories that started in one direction and reversed back to the original position (Figure \ref{3}A, Figure 3-Supplemental 1B), trajectories that stayed in one position (Figure \ref{3}B), trajectories that jumped between arms (Figure \ref{3}C-D, Figure3-Supplemental 1A, E, F), and trajectories that remained spatially incoherent throughout the SWR (Figure3-Supplemental1D). Overall, we were able to classify 89\% ( 23382/26160) of SWRs as containing at least one of the five dynamics. To verify that this depended on the spatial receptive fields of cells in the hippocampus, we resampled position with replacement for two epochs, breaking the relationship between spiking and spatial information. There were far fewer SWR classified when spatial information was removed (XX fewer SWRs, CI [], p=0.XX for epoch 1, p=0.XX for epoch 2, Figure 3-Supplemental 2).

@@

To verify that partitioning the data into spatially coherent and incoherent time segments by dynamic was meaningful, we quantified the cumulative spatial extent of the 95\% highest posterior density (HPD) values of the posterior linear position. Larger values of the cumulative HPD indicate the model is less certain about position, because the HPD values are spread out over more of the track at a given time point, whereas lower values indicate that the model is more certain about the estimate of position, because the HPD values are more concentrated and cover less of the track. For example, Figure \ref{4}A and B show two SWRs that were classified as having stationary and continuous dynamics, respectively. The bulk of the HPD values at each time step in these SWRs is concentrated on a small portion of the track and so the cumulative HPD is low throughout the SWRs. In contrast, the Figure \ref{4}C shows a SWR where the dynamics are fragmented and correspondingly, the HPD is much more spatially diffuse and the cumulative HPD is much higher. This can change over the time course of a SWR, as in Figure \ref{4}D, where the cumulative HPD is small for most of the SWR until the end, where the uncertainty of position becomes much higher. When we looked at the average cumulative HPD for each SWR by dynamic, we found a clear bi-modal split between spatially congruent dynamics and spatially incongruent dynamics (Figure \ref{4}E). For the spatially congruent dynamics, the cumulative HPD was much lower than the spatially incongruent dynamics (median XX cm, spatially congruent vs. median XX cm, spatially incongruent).

## Reminder: need colorbar and legend
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure5/Figure5_v2}
\caption{95 \% Highest Posterior Density of the Dynamics. \textbf{(A-D)}  Examples of the 95 \% Highest Posterior Density. Top panel. Probability of dynamic over time. Shading and labels indicates dynamic categories. 2nd panel. Posterior linear position over time. Red triangle indicates animal's position. 3rd panel. Black indicates the position bin is in the 95\% highest posterior density (HPD) values. 4th panel. Cumulative HPD. \textbf{(E)} Average Cumulative 95\% HPD for each dynamic category.}
\label{4}
\end{figure*}

\subsection*{Many replay trajectories have slower movement speeds}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure4/Figure4_v4}
\caption{
Analysis of SWRs from all animals. \textbf{(A)} UpSet plot---which is similar to a Venn diagram with more than 3 sets---of the most common sets of dynamics within each ripple. Each row represents a possible category present in the SWR. Each column represents a set of dynamics categories, where filled-in black dots with an edge between the dots indicates that multiple categories are present in the SWR (at temporally distinct times). The sets are ordered by how often they occurred as indicated by the bar plot above each category. The total number of each category is indicated by the rotated bar plot on the left. \textbf{(B)} Distribution of the duration of each dynamic category within a SWR. \textbf{(C)} The distance of the average latent position from the animal's position for each dynamic category within each SWR. \textbf{(D)} Position of stationary trajectories on the W-track at least 30 cm away from the animal's position. \textbf{(E)} Distribution of multiunit spike rates for each dynamic category within each SWR.
}
\label{5}
\end{figure*}

We next looked at which sets of dynamics are most common. Most of the SWRs (16454 of 23382 classified SWRs or 70\%) were well described by one dynamic. Surprisingly, the most common SWRs were ones that were categorized exclusively as stationary-continuous mixtures (9860 of 23382 classified SWRs or 42\%, Figure \ref{5}A). These trajectories had a slightly shorter duration to events with continuous dynamics (median duration: stationary-continuous-mixture 72 ms vs. continuous 95 ms, Figure \ref{5}B) and occurred at a similar average distance away from the animal (mean trajectory distance from animal's position: stationary-continuous-mixture 51 cm vs. continuous 42 cm, Figure \ref{5}C). This indicates that the trajectories of these SWRs, like ones that have faster continuous dynamics, do not represent the animal's position and move at a speed that is closer to the animal's actual movement speed when running, rather than many times the animal's movement speed.

The next most common category is those SWRs that were exclusively stationary (5149 / 23382 or 22\%). Unlike the stationary-continuous mixtures, most of these trajectories represented the animal's position (Figure \ref{5}C), but there were many that represented positions as far as 226 cm from the animal. Duration. We define the ones that are more than 30 cm away from the animal as "non-local stationary trajectories". There were 1585 of 23382 classified SWRs or 7\% non-local stationary trajectories and 731 of these non-local stationary trajectories were stationary throughout the duration of the SWR. These non-local stationary trajectories represented positions all over the track, but were most common at reward wells and choice points (Figure \ref{5}D), consistent with \cite{JaiDistincthippocampalcorticalmemory2017}. The average multiunit firing rate during stationary dynamics was not lower than the other dynamics (Figure \ref{5}E.  ## Need to fix multiunit panel

@@

\subsection*{Behavioral Task}
The behavioral task is a W-track spatial alternation task as presented in \cite{KarlssonAwakereplayremote2009}. For each day, animals alternated between 15 minute sessions in a rest box and 15 minutes in the W-track. The behavioral task is a W-track spatial alternation task as presented in \cite{KarlssonAwakereplayremote2009}. For each day, animals alternated between 15 minute sessions in a rest box and 15 minutes in the W-track. Only run epochs with at least 9 hippocampal principal cells that fired at least 100 spikes were included in the analysis.

Only run epochs with at least 9 hippocampal principal cells that fired at least 100 spikes were included in the analysis.

@@

>>>>

\subsection*{The model finds replays with non-constant speed dynamics}
After validating our model on a single SWR, we then applied our decoding algorithm to hippocampal recordings from 10 animals performing the W-track spatial alternation task (tetrodes = [10, 24], brain areas = [CA1, CA2, CA3]; some data previously used in \cite{KarlssonAwakereplayremote2009}). We observed many replays that were classified as continuous throughout the entirety of SWR and likely would have been identified by the typical Bayesian Decoder (Figure 2-Supplemental 2A-C). However, we also observed replays that did not have this structure, but could be interpreted in terms of our model. For example, we observed trajectories that started in one direction and reversed back to the original position (Figure \ref{3}A, Figure 3-Supplemental 1B), trajectories that stayed in one position (Figure \ref{3}B), trajectories that jumped between arms (Figure \ref{3}C-D, Figure3-Supplemental 1A, E, F), and trajectories that remained spatially incoherent throughout the SWR (Figure3-Supplemental1D). Overall, we were able to classify 89\% ( 23382 of 26160) of SWRs as containing at least one of the five dynamics. To check that this depended on the spatial receptive fields of cells in the hippocampus, we resampled position with replacement for two epochs, breaking the relationship between spiking and spatial information. There were far fewer SWR classified when spatial information was removed (99\% true classified, 9\% median resampled distribution, p=0.02 for epoch 1, p=0.02 for epoch 2, Figure 3-Supplemental 2).

@@

To verify that partitioning the data into spatially coherent and incoherent time segments by dynamic was meaningful, we quantified the cumulative spatial extent of the 95\% highest posterior density (HPD) values of the posterior linear position. Larger values of the cumulative HPD indicate the model is less certain about position, because the HPD values are spread out over more of the track at a given time point, whereas lower values indicate that the model is more certain about the estimate of position, because the HPD values are more concentrated and cover less of the track. For example, Figure \ref{4}A and B show two SWRs that were classified as having stationary and continuous dynamics, respectively. The bulk of the HPD values at each time step in these SWRs is concentrated on a small portion of the track and so the cumulative HPD is low throughout the SWRs. In contrast, the Figure \ref{4}C shows a SWR where the dynamics are fragmented and correspondingly, the HPD is much more spatially diffuse and the cumulative HPD is much higher. This can change over the time course of a SWR, as in Figure \ref{4}D, where the cumulative HPD is small for most of the SWR until the end, where the uncertainty of position becomes much higher. When we looked at the average cumulative HPD for each SWR by dynamic, we found a clear bi-modal split between spatially congruent dynamics and spatially incongruent dynamics (Figure \ref{4}E). For the spatially congruent dynamics, the cumulative HPD was much lower than the spatially incongruent dynamics (median 24 cm, spatially congruent vs. median 238 cm, spatially incongruent).

\begin{figure*}%[tb2]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure4/Figure4_v5}
\caption{95 \% Highest Posterior Density of the Dynamics. \textbf{(A-D)}  Examples of the 95 \% Highest Posterior Density. Top panel. Probability of dynamic over time. Shading and labels indicates dynamic categories. 2nd panel. Posterior linear position over time. Red triangle indicates animal's position. 3rd panel. Black indicates the position bin is in the 95\% highest posterior density (HPD) values. 4th panel. Cumulative HPD. \textbf{(E)} Average Cumulative 95\% HPD for each dynamic category.}
\label{4}
\end{figure*}

\subsection*{Many replay trajectories have slower movement speeds}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure5/Figure5_v3}
\caption{
Analysis of SWRs from all animals. \textbf{(A)} UpSet plot---which is similar to a Venn diagram with more than 3 sets---of the most common sets of dynamics within each ripple. Each row represents a possible category present in the SWR. Each column represents a set of dynamics categories, where filled-in black dots with an edge between the dots indicates that multiple categories are present in the SWR (at temporally distinct times). The sets are ordered by how often they occurred as indicated by the bar plot above each category. The total number of each category is indicated by the rotated bar plot on the left. \textbf{(B)} Distribution of the duration of each dynamic category within a SWR. \textbf{(C)} The distance of the average latent position from the animal's position for each dynamic category within each SWR. \textbf{(D)} Position of stationary trajectories on the W-track at least 30 cm away from the animal's position. \textbf{(E)} Distribution of multiunit spike rates for each dynamic category within each SWR.
}
\label{5}
\end{figure*}

We next looked at which sets of dynamics are most common. Most of the SWRs (16454 of 23382 classified SWRs or 70\%) were well described by a single dynamic. Surprisingly, the most common SWRs were ones that were categorized exclusively as stationary-continuous mixtures (9860 of 23382 classified SWRs or 42\%, Figure \ref{5}A). These trajectories had a slightly shorter duration to events with continuous dynamics (median duration: stationary-continuous-mixture 72 ms vs. continuous 95 ms, Figure \ref{5}B) and occurred at a similar average distance away from the animal (mean trajectory distance from animal's position: stationary-continuous-mixture 51 cm vs. continuous 42 cm, Figure \ref{5}C). This indicates that the trajectories of these SWRs, like ones that have faster continuous dynamics, do not represent the animal's position and move at a speed that is closer to the animal's actual movement speed when running, rather than many times the animal's movement speed.

The next most common category is those SWRs that were exclusively stationary (5149 of 23382 or 22\%). Unlike the stationary-continuous mixtures, most of these trajectories represented the animal's position (Figure \ref{5}C), but there were many that represented positions as far as 226 cm from the animal. We define the ones that are more than 30 cm away from the animal as "non-local stationary trajectories". There were 1585 of 23382 classified SWRs or 7\% non-local stationary trajectories and 731 of these non-local stationary trajectories were stationary throughout the duration of the SWR. These non-local stationary trajectories represented positions all over the track, but were most common at reward wells and choice points (Figure \ref{5}D), consistent with \cite{JaiDistincthippocampalcorticalmemory2017}. The average multiunit firing rate during stationary dynamics was not lower than the other dynamics (Figure \ref{5}E.

@@

\subsection*{Behavioral Task}
The behavioral task is a W-track spatial alternation task as presented in \cite{KarlssonAwakereplayremote2009}. In brief, each day, animals alternated between 15 minute rest sessions in a rest box and 15 minutes run session in the W-track. On the W-track, animals had to run: inbound and outbound. On outbound runs, the Only run epochs with at least 9 hippocampal principal cells that fired at least 100 spikes were included in the analysis.

@@

\subsection{Highest Posterior Density}
